\vspace{-1.5em} \bf Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-06-08 paper.tex

@@

The properties of the mutations inferred in the Wide and Long
ARGs are summarised in Figure~\ref{fig:mutational_spectra}B. In both
cases we have a large number of mutations, and a majority of these
(Wide ARG: 62.55\%, Long ARG: 72.23\%) are private to a single
sample, i.e.\ on terminal branches.
Although the average number
of mutations per sample is small
(Wide ARG: 0.77, Long ARG: 1.38; Table~\ref{tab:args}),
the average number per site
is large in both ARGs (Wide ARG: 41.23; Long ARG:36.10; Table~\ref{tab:args}).
However, a lower median count (Wide ARG: 14; Long ARG:13),
suggests that these high mutation counts are
partly driven by some hypermutable sites (e.g., site 28,271 has over 7,000
mutations in the Wide ARG; Figure~\ref{fig:breakpoint-distribution}), which
may be artefactual.
A significant issue with the current version of \texttt{sc2ts} is an excess
of inferred reversion mutations, with 5.99\% of all mutations in the Wide
ARG (Long ARG: 4.54\%) reverting the state change of the immediately
ancestral mutation (see Section~\ref{sec:args} for precise definitions).
The majority of these mutations are likely to be
artefactual, and a key goal of future development is to improve
our parsimony heuristics (Section~\ref{sec:parsimony-heuristics}) to
address this issue.

@@

\subsection{Recombination breakpoint intervals}
\label{sec:breakpoint_intervals}
It is rarely possible to be precise about the position on the genome at which
a recombinant sequence switches from inheriting from one parent to another.
Even if we observe the recombinant sequence before subsequent
divergence occurs (during onward transmission), if the two parent sequences
are similar there is no way to
accurately identify the breakpoint.
Here we define the
interval around a particular breakpoint as the genome coordinates over which the
sequences for the left and right parent nodes are identical. The right-hand
extreme of the breakpoint interval is chosen by the LS HMM Viterbi algorithm
(Section~\ref{sec:ls}), and the left endpoint is then derived by directly
comparing the parent sequences. See Section~\ref{sec:treatment_recombinants} for
a precise definition of breakpoints and their intervals.

@@

Although we emphasise that further work is required to filter out spuriously inferred recombination events,
we can draw some preliminary conclusions from Figure~\ref{fig:breakpoint-distribution}.
The number of intervals spanning a site (orange curve) is lower at the ends of the genome,
as expected due to the lack of information about recombination with few flanking sites.
In addition, the number of intervals spanning a site often
drops near the beginning of a gene.
This is particularly apparent at the ORF8/N gene interface,
with the N gene containing fewer potential recombination breakpoints
than other genes, in agreement with the results of \cite{Turakhia2022-it}.
Noticeable declines in the number of spanning intervals are also
seen near the beginning of S, M, ORF7a, and ORF8.
These declines are sometimes associated with hypermutated sites (e.g., 27,384 and 28,271
near the beginning of ORF7a and N, respectively), as expected because sites that
undergo mutation at high rates are more likely to differ between the
parents of a recombinant and so provide information about the location of a breakpoint.
This pattern may, however, also be an artefact of sequencing errors causing sites
to appear different between the parents when they are not
(see the discussion of potential errors at site 28,271 below).
In other cases, however, the drop in intervals does not appear to
coincide with hypermutability and may reflect shifts in the actual
rate of recombination between genes.
Indeed, template switching is known to occur at hotspots,
which often involve transcription-regulatory sequences preceding genes
in SARS-CoV-2~\citep{Yang2021-characterizing}
The rate of recombination may also depend on the relative abundance
of different subgenomic RNA intermediates that span different genes,
affecting the availability of templates and influencing the rate of
homologous recombination~\citep{Kim2020-gt,Zou2021-sars}
[Sally: KATHERINE, WHAT DO YOU THINK?  These not-full-length subgenomic RNA intermediates may provide a lot of homologous templates for some genes more than other (e.g., fig 3B here)]

It is important to note that the precise endpoints of intervals can be
somewhat arbitrary, since they are defined by exact comparisons
of a recombinant's parents.
Thus, breakpoint intervals will tend to be truncated at sites that are
hypermutable, either due to increased information about parentage
or spurious inferences caused by sequencing errors.
It is therefore helpful to compare the number of intersecting intervals
with the number of mutations per site in Figure~\ref{fig:breakpoint-distribution}.
For instance, position 28,271 (located in the ORF8-N intergenic
region) has the largest number of mutations and appears as an endpoint
of 66 intervals.
The 7,572 mutations (including 5,782 insertions and 1,605 deletions)
at this site are an indicator that this
homopolymeric region may be prone to sequencing errors
and should be included in the list of ``problematic sites'' that
are excluded from analysis (Section~\ref{sec:data_preprocessing}).
On the other hand, 58 of the breakpoint intervals have endpoints
within one base of site 27,972, which has undergone 770 mutation events
(including 425 C$>$T, 314 T$>$C). The C$>$T mutation has the effect of
truncating ORF8, and it has been posited that the truncation is neutral
or advantageous for transmission, and disadvantageous within-host
\citep{Jungreis2021-dh}, suggesting that the high rate of recurrent mutation
at this site may be due to selection.

@@

The most recent recombination node for 79 samples (XN: 53, XZ: 16, XAJ: 6, XE:
1, XAD: 1, XAH: 1, XAK: 1) is the same. This node, introduced as a
result of adding strain
Denmark/DCGC-281594/2021, is ancestral to more than 127K samples,
and is highly likely to be a spuriously inferred recombination event.
Understanding the origin of this
and other similar high-impact, but likely spurious recombination
events is an important aspect of future work.
For simplicity, we exclude these samples from further analyses in
this section.

@@

For XB, all samples trace back to an originating node which is the product of a
recombination between a B.1 sample (specifically the strain
England/CAMB-7B47D/2020, which contributed the majority of
the genome from the start up to a rightmost position of 23604bp), and two
UPGMA nodes with imputed Pango lineages B.1.627 (up to a rightmost position of
27389bp) and B.1.36.8 (the remaining fragment of the genome). Figure~\ref{fig:complex_origins_graph}
shows that in the Long ARG the equivalent recombination node has only 2 parents,
with no involvement of  B.1.36.8; it is possible that the third parent in
the Wide ARG is artefactual. As in the Long ARG, additional
non-X-designated samples such as B.1.634 also descend from this recombination,
and there are also a small number of nested recombination nodes. However, all but
one of these nested recombinations are unimportant, being ancestral to negligible fractions
of the Nextclade-designated XB samples. The one exception accounts for about 17\% of the
designated XB nodes and involves a recombination between descendants of the
originating XB recombination. More specifically, strain
USA/TX-HMH-MCoV-43092/2021 is inferred to be a recombinant
between an XB grandparent and its XB grandchild, involving an intermediate UPGMA node.
It seems likely to be artefactual recombination event
caused by undersampling, but could reflect true 
recombination between closely related lineages.

@@

\subsection{False positive recombinants}
\label{sec:false_positives}
The Long and Wide ARGs both contain several recombination nodes that are
ancestral to large numbers of samples, and are almost certainly
artefactual. Table~\ref{tab:false_positive} shows details for four recombinants
in the Long ARG that are likely false positives, and have come to be
the ancestors of a large number of samples. These are the top-four recombinants
from the Long ARG in terms of numbers of descendant samples
(the fifth-largest has substantially fewer, at 14,869 descendant samples),
labelled A--D. These false positives are associated with the origins
of large outbreaks. For each row, we show the sample's Pango lineage
designation and the rank of that sample in the overall list of
samples with that designation. Thus, rows C and D are the first
samples seen in the Long ARG from the BA.2.9 and BA.3 lineages,
and A and B are the third and eighth samples from the B.1.617.2
lineage (Delta VoC), respectively. Also shown is the overall ``cost'' of the Viterbi
solution computed by the LS HMM (Section~\ref{sec:ls}), which
is $3 \times$ number of recombinations + number of additional mutations
(for a mismatch ratio of 3; Section~\ref{sec:ls}).
The column shows that each of these samples was a large
``distance'' from the current ARG when it was added. The mean
HMM cost over all 2,078 recombinants in the Long ARG is 8.39
(median: 7), and the mean cost for the  763 non-singleton
recombinants is 8.5 (median: 7). Thus, the recombinants in
Table~\ref{tab:false_positive} are in the tail of this distribution.

@@

Occasional evolutionary leaps, in which a large number of mutations
are acquired in sudden jumps
is a signature feature of SARS-CoV-2~\citep{Corey2021-sars,Nielsen2023-host}.
Such ``saltations'' naturally present challenges to \texttt{sc2ts} and the
current HMM parameterisation of three mutations per recombination (mismatch
ratio). The first few sequences from these new lineages will match in to
the existing ARG with very low likelihood, and the HMM will therefore
search for ways to reduce the number of mutations required by recombining
in segments. If this is combined with data quality issues and
``time travellers'' (Section~\ref{sec:filtering_time_travellers}),
the early stages of the VoC emergence can be quite muddled.
This appears to be the case for the Delta VoC,
corresponding to rows A and B of Table~\ref{tab:false_positive},
which emerged and quickly rose to high prevalence in early-mid 2021,
carrying 30 characteristic mutations compared to the reference sequence 
\citep{McCrone2022-context}, including nine mutations in the S gene 
not seen in earlier VoCs.
Similar patterns around the emergence of Delta are seen in the Wide ARG
(not shown).

@@

% PARA 3. Handling the data is hard, so methods for making this easier
% are important. Classical phylo approaches don't scale to millions.
% The ability to quickly develop new analyses, at scale, will be crucial.
% Sc2ts is largely composed of other parts being reused, with mainly being
% concerned with masking input alignments. We wrote no C at all. This kind
% of code reuse and ease of analysis is a major bonus.
The vast volume of whole genome sequence data generated during the pandemic
has presented classical phylogenetic methods and software with major
difficulties~\citep{Hodcroft2021-wt}.
Standard interchange formats such as FASTA, Newick and VCF were simply not designed
to deal with millions of samples, and their limitations have come sharply
into focus~\citep{Turakhia2021-ur,de2023maximum}.
Replacements that can scale to millions of genomes have had to be developed
at speed, usually focusing on compiled programming languages to maximise
performance. Here, however, we have developed a new method based on
an existing data structure and library infrastructure, designed
from the beginning to scale to millions of
samples~\citep{Kelleher2016-wk,Kelleher2019-ba}.
The \texttt{sc2ts} package is written entirely in Python
and by reusing existing high-performance components can infer
recombinant viral ancestry at unprecedented scale.
Similarly, all of the analyses shown here are written in Python,
using the \texttt{tskit} API, mostly running in a few seconds
on standard laptop computers (see the Data Availability
section for details of the corresponding Jupyter notebooks).
Retooling methods to scale rapidly and to encode recombination
promises to improve tracking during this and any future pandemic.

@@

The generative process of the LS model is summarised in
Figure~\ref{fig:ls_diagram}. Here, a transition matrix, $Q$, governs the
process of switching (recombining) between members of the reference panel (the
hidden states). An emission matrix, $E$, allows for differences between the
focal sequence and the hidden state from which it is copied (due to mutation as
well as sequencing error).
Both $E$ and $Q$ may be a function of the reference panel members,
but transitions are generally assumed to be independent of the
hidden states (Figure~\ref{fig:ls_diagram}, pink panel).
This assumption dramatically increases performance as the state space drops to two states (i.e., switching or not switching).
Emissions may also be a function of the nucleotide states, but in our
case, we assume that all alleles ($A,C,G,T,-$) can be mutated to with equal probability $\mu_\ell/5$. This is reasonable for rapidly evolving pathogens, but we note that setting the number of alleles at site $\ell$ ($a_\ell$) to the set of observed alleles across all analysed samples ($\mu_\ell/(a_\ell-1)$) will often be more appropriate
 (Figure~\ref{fig:ls_diagram}, blue panel).
We use the Viterbi algorithm~\citep{Viterbi1967-ol}
to find the most likely copying path, given a set of reference sequences, $Q$, and $E$.
Throughout, we refer to the probabilities of mismatching to a
member of the reference panel at site $\ell$ as $\mu_\ell$,
and the probability of recombining between two members of the reference panel between site $\ell-1$ and site $\ell$ as $r_\ell$. For convenience, $\mu_\ell$ is commonly referred to as the `mutation probability', but we note that this probability of
mismatch also encompasses various other error modes that result in a mismatch
(such as sequencing and alignment errors). Note also that it is these probabilities,
not the rates of mutation and recombination, that are required to fully define
the HMM; see~\cite{Donnelly2010-coalescent} for a discussion of how these
parameters relate to the coalescent process.

@@

The probabilities of these competing processes of mismatch and recombination are
usually controlled by the site-specific parameters $\mu_\ell$ and $r_\ell$,
respectively. For this work we used a slightly different formulation, which
uses one parameter, the mismatch ratio (MMR), to control the relative
importance of mutation and recombination in the HMM. Specifically, an MMR value
of $k$ will prefer $k$ mismatches (mutations) to a single recombination that
 results in copying from a template with no mismatches. To map between
recombination probability and mutation probability for a particular mismatch
ratio, we simply consider the two paths that we wish to be equally likely, and
rearrange for the mutation or recombination probability. Consider the simple
case where we assume $\mu_l=\mu$ and $r_l=r$. Without loss of generality,
consider a region of length $m$. Up until this region, two paths are equally
likely, so we can re-scale by the likelihood of observing the focal sequence up
to the site before the region starts, $c$. For an MMR of $k$, we only need to
consider two of the $n$ members of the reference panel.
One, $\mathcal{P}_1$,
for which there are no mismatches, but we need to recombine in, and a second,
$\mathcal{P}_2$, for which there are $k$ (randomly chosen) mismatches in the
region.
Now, follow the probability of tracking along each of those paths:
\begin{align*}
\mathbb{P}[\mathcal{P}_1] &= \frac{cr}{n}
\alpha^{m-1}\left(1-\mu\right)^m
&\text{recombine to a template with no mismatches,}\\
\mathbb{P}[\mathcal{P}_2] &= c
\alpha^{m}
% \alpha \alpha^{m-1}
\left(1-\mu\right)^{m-k}\mu^k &\text{stay in a template with $k$ mismatches,}
\end{align*}
% where $r$ is the recombination probability between adjacent
% variants, $\mu$ is the mismatch probability between a target sequence at a
% variant and a particular member of the reference panel,
where $\alpha = (1 - r + r / n) $ and $c$ is the
likelihood of the path up to this point, which are assumed to be
equal by construction.
We then
set these path probabilities to be equal, and rearrange to relate $r$ and $\mu$ to one another:
\begin{align*} r = \frac{n\mu^k}{\left(1-\mu\right)^k + \left(n-1\right)\mu^k},
\quad\quad \mu = \frac{1}{\sqrt[k]{\frac{n}{r} - (n-1)} + 1}. \end{align*}
Thus, for lower MMR values ($k$ here),
recombinations are favoured over mutations
than a single ancestral genome.
We use an MMR value of $k=3$ in this work,
as it provided results that tallied well with empirical observations.
Exploring different mismatch ratios and more sophisticated parameterisations
of the HMM are important avenues for future work.

A varying degree of missingness is present in the genetic data, that is, for a
given sample, a subset of nucleotides may be coded as `unknown' at a set of positions.
As a result, we had to consider missingness in our implementation of the LS
Viterbi algorithm. One approach would be to `impute' all missing data
beforehand using external software to hard-call or define a distribution over
the possible missing nucleotides, and use that in the determination of path
probabilities. Another, which we carry out within \texttt{sc2ts}, is to not attempt to estimate the true nucleotide, but
rather to consider such sites as uninformative to the path probability, and
impute missing data within that path on the fly, placing them within the emission probability transition from unknown to
the emission probability from unknown to
to any other state as 1.

@@

There are well-known issues with using such a simple algorithm for inferring
evolutionary relationships~\citep{Felsenstein2004-inferring}.
Table~\ref{tab:args} show that
this within-cluster tree building has a significant influence on the
overall ARG topology
and therefore applying more sophisticated
tree building methods which keep track of the required mutations
(rather then inferring post-hoc by parsimony), is a likely avenue
for improvements in overall inference quality.

\subsection{Parsimony-increasing heuristics}
\label{sec:parsimony-heuristics}
Attaching trees built from the clusters of samples that copy from
a particular node (or path of nodes for recombinants,
see Section~\ref{sec:treatment_recombinants}) under the
HMM is an inherently greedy strategy,
and can produce inferences that are clearly unparsimonious.
The final step in adding a daily batch of samples to the ARG
is therefore to perform some local updates that target specific
types of parsimony violations in the just-updated regions of the
ARG. There are currently two parsimony-increasing operations
applied, which we refer to as ``mutation collapsing'' and ``reversion
pushing'' (Figure~\ref{fig:overview_sc2ts}D, E).

>>>>

% sets colours for notes to each other in the text
\usepackage{xcolor}
\newcommand{\sally}[1]{\textcolor{red}{#1}}
\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\green}[1]{\textcolor{green}{#1}}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}

@@

The properties of the mutations inferred in the Wide and Long
ARGs are summarised in Figure~\ref{fig:mutational_spectra}B. In both
cases we have a large number of mutations, and a majority of these
(Wide ARG: 62.55\%, Long ARG: 72.23\%) are private to a single
sample, i.e.\ on terminal branches.
Although the average number
of mutations per sample is small
(Wide ARG: 0.77, Long ARG: 1.38; Table~\ref{tab:args}),
the average number per site
is large in both ARGs (Wide ARG: 41.23; Long ARG: 36.10; Table~\ref{tab:args}).
However, a lower median count (Wide ARG: 14; Long ARG: 13),
suggests that these high mutation counts are
partly driven by some hypermutable sites (e.g., site 28,271 has over 7,000
mutations in the Wide ARG; Figure~\ref{fig:breakpoint-distribution}), which
may be artefactual.
A significant issue with the current version of \texttt{sc2ts} is an excess
of inferred reversion mutations, with 5.99\% of all mutations in the Wide
ARG (Long ARG: 4.54\%) reverting the state change of the immediately
ancestral mutation (see Section~\ref{sec:args} for precise definitions). \sally{[What confuses me is that I would think that none of these reversions would remain after the reversion push step of Figure 1D.  Is this because ``immediately ancestral'' is incorrect or because the problem is that some of these immediate reversions are created deeper in the tree than the newly added samples (i.e., uncorrected by the reversion push). If the former, maybe say ``reverting the state change of a recent ancestral mutation'', if the latter, maybe say  ``(see Section~\ref{sec:args} for precise definitions, note that the reversion push algorithm of Figure 1D only eliminates reversions in the newly added portion of the tree and not reversions around earlier nodes created by this algorithm)''.]}
The majority of these mutations are likely to be
artefactual, and a key goal of future development is to improve
our parsimony heuristics (Section~\ref{sec:parsimony-heuristics}) to
address this issue.

@@

\subsection{Recombination breakpoint intervals}
\label{sec:breakpoint_intervals}
It is rarely possible to be precise about the position on the genome at which
a recombinant sequence switches from inheriting from one parent to another.
Even if we observe the recombinant sequence before subsequent
divergence occurs (during onward transmission), there is no way to identify the exact breakpoint if the two parent sequences
are similar.
Here we define the
interval \sally{within which a particular breakpoint may have occurred} as the genome coordinates over which the
sequences for the left and right parent nodes are identical. The right-hand
extreme of the breakpoint interval is chosen by the LS HMM Viterbi algorithm
(Section~\ref{sec:ls}), and the left endpoint is then derived by directly
comparing the parent sequences. See Section~\ref{sec:treatment_recombinants} for
a precise definition of breakpoints and their intervals.

@@

Although we emphasise that further work is required to filter out spuriously inferred recombination events,
we can draw some preliminary conclusions from Figure~\ref{fig:breakpoint-distribution}.
The number of intervals spanning a site (orange curve) is lower at the ends of the genome,
as expected due to the lack of information about recombination with few flanking sites.
In addition, the number of intervals spanning a site often
drops near the beginning of a gene.
This is particularly apparent at the ORF8/N gene interface,
with the N gene containing fewer potential recombination breakpoints
than other genes, in agreement with the results of \cite{Turakhia2022-it}.
Noticeable declines in the number of spanning intervals are also
seen near the beginning of S, M, ORF7a, and ORF8.
These declines are sometimes associated with hypermutated sites (e.g., 27,384 and 28,271
near the beginning of ORF7a and N, respectively), as expected because sites that
undergo mutation at high rates are more likely to differ between the
parents of a recombinant and so provide information about the location of a breakpoint.
This pattern may, however, also be an artefact of sequencing errors causing sites
to appear different between the parents when they are not
(see the discussion of potential errors at site 28,271 below).
In other cases, however, the drop in intervals does not appear to
coincide with hypermutability and may reflect shifts in the actual
rate of recombination between genes.
Indeed, template switching is known to occur at hotspots,
which often involve transcription-regulatory sequences preceding genes
in SARS-CoV-2~\citep{Yang2021-characterizing}
The rate of recombination may also depend on the relative abundance
of different subgenomic RNA intermediates that span different genes,
affecting the availability of templates and influencing the rate of
homologous recombination~\citep{Kim2020-gt,Zou2021-sars}
\sally{[Sally: KATHERINE, WHAT DO YOU THINK?  These not-full-length subgenomic RNA intermediates may provide a lot of homologous templates for some genes more than other (e.g., fig 3B here).]}

It is important to note that the precise endpoints of intervals can be
somewhat arbitrary, because they are defined by \sally{the sequence differences that happen to be present in the} recombinant's parents.
Thus, breakpoint intervals will tend to be truncated at sites that are
hypermutable, either due to increased information about parentage
or spurious inferences caused by sequencing errors.
It is therefore helpful to compare the number of intersecting intervals
with the number of mutations per site (Figure~\ref{fig:breakpoint-distribution}).
For instance, position 28,271 (located in the ORF8-N intergenic
region) has the largest number of mutations and appears as an endpoint
of 66 intervals.
The 7,572 mutations (including 5,782 insertions and 1,605 deletions)
at this site are an indicator that this
homopolymeric region may be prone to sequencing errors
and \sally{potentially} should be included in the list of ``problematic sites'' that
are excluded from analysis (Section~\ref{sec:data_preprocessing}).
On the other hand, 58 of the breakpoint intervals have endpoints
within one base of site 27,972, which has undergone 770 mutation events
(including 425 C$>$T, 314 T$>$C). The C$>$T mutation has the effect of
truncating ORF8, and it has been posited that the truncation is neutral
or advantageous for transmission, and disadvantageous within-host
\citep{Jungreis2021-dh}, suggesting that the high rate of recurrent mutation
at this site may be due to selection.

@@

\sally{[Delete this paragraph now? The new section 4.8 gets at the real issue, that the multi-step mutations common in new VoC generate false positive signals of recombination.]} The most recent recombination node for 79 samples (XN: 53, XZ: 16, XAJ: 6, XE:
1, XAD: 1, XAH: 1, XAK: 1) is the same. This node, introduced as a
result of adding strain
Denmark/DCGC-281594/2021 (BA.2.9), is ancestral to more than 127K samples, 
and is highly likely to be a spuriously inferred recombination event. \sally{[Replace with ``...is ancestral to more than 127K samples and XX immediate descendants. While such a hyper-connected node may reflect a large outbreak with little phylogenetic resolution, it may also be spurious, e.g., caused by the attraction of sequences with similar constellations of mutations or another phylogenetic artefact.'']}
Understanding the origin of this
and other similar high-impact, but potentially spurious, recombination
events is an important aspect of future work.
For simplicity, we exclude these samples from further analyses in
this section.

@@

For XB, all samples trace back to an originating node which is the product of a
recombination between a B.1 sample (specifically the strain
England/CAMB-7B47D/2020, which contributed the majority of
the genome from the start up to a rightmost position of 23604bp), and two
UPGMA nodes with imputed Pango lineages B.1.627 (up to a rightmost position of
27389bp) and B.1.36.8 (the remaining fragment of the genome). Figure~\ref{fig:complex_origins_graph}
shows that in the Long ARG the equivalent recombination node has only 2 parents,
with no involvement of  B.1.36.8; it is possible that the third parent in
the Wide ARG is artefactual. As in the Long ARG, additional
non-X-designated samples such as B.1.634 also descend from this recombination,
and there are also a small number of nested recombination nodes. However, all but
one of these nested recombinations are unimportant, being ancestral to negligible fractions
of the Nextclade-designated XB samples. The one exception accounts for about 17\% of the
designated XB nodes and involves a recombination between descendants of the
originating XB recombination. More specifically, strain
USA/TX-HMH-MCoV-43092/2021 is inferred to be a recombinant
between an XB grandparent and its XB grandchild, involving an intermediate UPGMA node.
It seems likely to be an artefactual recombination event caused by undersampling, but it could also reflect true recombination between closely related lineages.

@@

\subsection{False positive recombinants}
\label{sec:false_positives}
The Long and Wide ARGs both contain several recombination nodes that are
ancestral to a large number of samples\sally{. These inferences of recombination may well be artefacts caused by the appearance of new variants of concern carrying more than the expected number of mutations \cite{otto2021origins}}. Table~\ref{tab:false_positive} shows details for four recombinants
in the Long ARG that are likely false positives and have come to be
the ancestors of a large number of samples. These are the top-four recombinants
from the Long ARG in terms of numbers of descendant samples
(the fifth-largest has substantially fewer, at 14,869 descendant samples),
labelled A--D. 
\sally{[DELETED SENTENCE]}%These false positives are associated with the origins of large outbreaks. 
For each row, we show the sample's Pango lineage
designation and the rank of that sample in the overall list of
samples with that designation. Thus, rows C and D are the first
samples seen in the Long ARG from the BA.2.9 and BA.3 lineages,
and A and B are the third and eighth samples from the B.1.617.2
lineage (Delta VoC), respectively. Also shown is the overall ``cost'' of the Viterbi
solution computed by the LS HMM (Section~\ref{sec:ls}), which
is $3 \times$ number of recombinations + number of additional mutations
(for a mismatch ratio of 3; Section~\ref{sec:ls}).
The column shows that each of these samples was a large
``distance'' from the current ARG when it was added. The mean
HMM cost over all 2,078 recombinants in the Long ARG is 8.39
(median: 7), and the mean cost for the  763 non-singleton
recombinants is 8.5 (median: 7). Thus, the recombinants in
Table~\ref{tab:false_positive} are in the tail of this distribution.

@@

Occasional evolutionary leaps, in which a large number of mutations
are acquired in sudden jumps,
is a signature feature of SARS-CoV-2~\citep{Corey2021-sars,otto2021origins,Nielsen2023-host}.
Such ``saltations'' naturally present challenges to \texttt{sc2ts} and the
current HMM parameterisation of three mutations per recombination (mismatch
ratio). The first few sequences from these new lineages will \sally{be a poor match} to
the existing ARG, and the HMM will therefore
search for ways to reduce the number of mutations required by recombining
segments. 
\sally{[DELETED SENTENCE (TRUE WHETHER OR NOT QUALITY ISSUES OCCUR).] }%If this is combined with data quality issues and ``time travellers'' (Section~\ref{sec:filtering_time_travellers}), the early stages of the VoC emergence can be quite muddled.
This appears to be the case for the Delta VoC,
corresponding to rows A and B of Table~\ref{tab:false_positive},
which emerged and quickly rose to high prevalence in early-mid 2021,
carrying 30 characteristic mutations compared to the reference sequence 
\citep{McCrone2022-context}, including nine mutations in the S gene 
not seen in earlier VoCs.
Similar patterns around the emergence of Delta are seen in the Wide ARG
(not shown).

@@

% PARA 3. Handling the data is hard, so methods for making this easier
% are important. Classical phylo approaches don't scale to millions.
% The ability to quickly develop new analyses, at scale, will be crucial.
% Sc2ts is largely composed of other parts being reused, with mainly being
% concerned with masking input alignments. We wrote no C at all. This kind
% of code reuse and ease of analysis is a major bonus.
The vast volume of whole genome sequence data generated during the pandemic
has presented classical phylogenetic methods and software with major
difficulties~\citep{Hodcroft2021-wt}.
Standard interchange formats such as FASTA, Newick and VCF were simply not designed
to deal with millions of samples, and their limitations have come sharply
into focus~\citep{Turakhia2021-ur,de2023maximum}.
Replacements that can scale to millions of genomes have had to be developed
at speed, usually focusing on compiled programming languages to maximise
performance. Here, however, we have developed a new method based on
an existing data structure and library infrastructure, designed
from the beginning to scale to millions of
samples~\citep{Kelleher2016-wk,Kelleher2019-ba}.
The \texttt{sc2ts} package is written entirely in Python
and by reusing existing high-performance components can infer
recombinant viral ancestry at unprecedented scale.
Similarly, all of the analyses shown here are written in Python,
using the \texttt{tskit} API, mostly running in a few seconds
on standard laptop computers (see the Data Availability
section for details of the corresponding Jupyter notebooks).
Retooling methods to scale up rapidly with expanding data and to encode recombination
promises to improve tracking during this and any future pandemic.

@@

The generative process of the LS model is summarised in
Figure~\ref{fig:ls_diagram}. Here, a transition matrix, $Q$, governs the
process of switching (recombining) between members of the reference panel (the
hidden states). An emission matrix, $E$, allows for differences between the
focal sequence and the hidden state from which it is copied (due to mutation as
well as sequencing error).
Both $E$ and $Q$ may be a function of the reference panel members,
but transitions are generally assumed to be independent of the
hidden states (Figure~\ref{fig:ls_diagram}, pink panel).
This assumption dramatically increases performance as the state space drops to two states (i.e., switching or not switching).
Emissions may also be a function of the nucleotide states, but in our RNA virus case, we assume that \sally{mutations occur to all possible alternate alleles} ($A,C,G,U,-$) with equal probability $\mu_\ell/5$. \sally{[Why is this 5?  I wouldn't have counted mutations to the same allele as a mutation.]} This is reasonable for rapidly evolving pathogens, but we note that setting the number of alleles at site $\ell$ ($a_\ell$) to the set of observed alleles across all analysed samples ($\mu_\ell/(a_\ell-1)$) will often be more appropriate
 (Figure~\ref{fig:ls_diagram}, blue panel).
We use the Viterbi algorithm~\citep{Viterbi1967-ol}
to find the most likely copying path, given a set of reference sequences, $Q$, and $E$.
Throughout, we refer to the probabilities of mismatching to a
member of the reference panel at site $\ell$ as $\mu_\ell$,
and the probability of recombining between two members of the reference panel between site $\ell-1$ and site $\ell$ as $r_\ell$. For convenience, $\mu_\ell$ is commonly referred to as the `mutation probability', but we note that this probability of
mismatch also encompasses various other error modes that result in a mismatch
(such as sequencing and alignment errors). Note also that it is these probabilities,
not the rates of mutation and recombination, that are required to fully define
the HMM; see~\cite{Donnelly2010-coalescent} for a discussion of how these
parameters relate to the coalescent process.

@@

The probabilities of these competing processes of mismatch and recombination are
usually controlled by the site-specific parameters $\mu_\ell$ and $r_\ell$,
respectively. For this work we used a slightly different formulation, which
uses one parameter, the mismatch ratio (MMR), to control the relative
importance of mutation and recombination in the HMM. Specifically, an MMR value
of $k$ will prefer $k$ mismatches (mutations) to a single recombination that
 results in copying from a template with no mismatches. To map between
recombination probability and mutation probability for a particular mismatch
ratio, we simply consider the two paths that we wish to be equally likely and
rearrange for the mutation or recombination probability. Consider the simple
case where we assume $\mu_l=\mu$ and $r_l=r$. Without loss of generality,
consider a region of length $m$. Up until this region, two paths are equally
likely, so we can re-scale by the likelihood of observing the focal sequence up
to the site before the region starts, $c$. For an MMR of $k$, we only need to
consider two of the $n$ members of the reference panel.
One, $\mathcal{P}_1$,
for which there are no mismatches, but we need to recombine in, and a second,
$\mathcal{P}_2$, for which there are $k$ (randomly chosen) mismatches in the
region.
Now, follow the probability of tracking along each of those paths:
\begin{align*}
\mathbb{P}[\mathcal{P}_1] &= \frac{cr}{n}
\alpha^{m-1}\left(1-\mu\right)^m
&\text{recombine to a template with no mismatches,}\\
\mathbb{P}[\mathcal{P}_2] &= c
\alpha^{m}
% \alpha \alpha^{m-1}
\left(1-\mu\right)^{m-k}\mu^k &\text{stay in a template with $k$ mismatches,}
\end{align*}
% where $r$ is the recombination probability between adjacent
% variants, $\mu$ is the mismatch probability between a target sequence at a
% variant and a particular member of the reference panel,
where \sally{$\alpha = (1 - r(n-1)/n) $ represents the probability of a sequence not recombining at a given position with any of the other $n-1$ members of the reference panel [REVISION OK?]} and $c$ is the
likelihood of the path up to this point, which are assumed to be
equal by construction.
We then
set these path probabilities to be equal, and rearrange to relate $r$ and $\mu$ to one another:
\begin{align*} r = \frac{n\mu^k}{\left(1-\mu\right)^k + \left(n-1\right)\mu^k},
\quad\quad \mu = \frac{1}{\sqrt[k]{\frac{n}{r} - (n-1)} + 1}. \end{align*}
Thus, for lower MMR values ($k$ here),
\sally{recombination is increasingly favoured over multiple mutations in a specific ancestral genome}.
%We use an MMR value of $k=3$ in this work, as it provided results that tallied well with empirical observations.
\sally{We use an MMR value of $k=3$ in this work,
because of the relatively high rate of recombination relative to mutation typical of coronaviruses \cite{amoutzias2022remarkable}.}
Exploring different mismatch ratios and more sophisticated parameterisations
of the HMM are important avenues for future work.

A varying degree of missingness is present in the genetic data, that is, for a
given sample, a subset of nucleotides may be coded as `unknown' at a set of positions.
As a result, we had to consider missingness in our implementation of the LS
Viterbi algorithm. One approach would be to `impute' all missing data
beforehand using external software to hard-call or define a distribution over
the possible missing nucleotides and use that in the determination of path
probabilities. Another, which we carry out within \texttt{sc2ts}, is to not attempt to estimate the true nucleotide but
rather to consider such sites as uninformative to the path probability, and
impute missing data within that path on the fly, placing them within the emission probability transition from unknown to
the emission probability from unknown to
to any other state as 1. \sally{[Repeated clause in the last sentence?  Also, I'm a bit unclear if this is enough.  Let's say that the nodes were G->Unknown->G.  The latter is given a probability of one, but is the former G->Unknown also allowed with probability one?  Note, however, that G->G shouldn't occur with probability one but Exp[-mu/t]]}

@@

There are well-known issues with using such a simple algorithm for inferring
evolutionary relationships~\citep{Felsenstein2004-inferring}.
Table~\ref{tab:args} shows that
this within-cluster tree building has a significant influence on the
overall ARG topology,
and therefore applying more sophisticated
tree building methods that keep track of the required mutations
(rather then inferring post-hoc by parsimony) is a likely avenue
for improvements in overall inference quality.

\subsection{Parsimony-increasing heuristics}
\label{sec:parsimony-heuristics}
Attaching trees built from the clusters of samples that copy from
a particular node (or path of nodes for recombinants,
see Section~\ref{sec:treatment_recombinants}) under the
HMM is an inherently greedy strategy
and can produce inferences that are clearly unparsimonious.
The final step in adding a daily batch of samples to the ARG
is therefore to perform some local updates that target specific
types of parsimony violations in the just-updated regions of the
ARG. There are currently two parsimony-increasing operations
applied, which we refer to as ``mutation collapsing'' and ``reversion
pushing'' (Figure~\ref{fig:overview_sc2ts}D, E).

Given a newly attached node, mutation collapsing inspects its siblings from
previous sample days to check if any of them share (a subset of) the mutations that
it carries. If so, we increase the overall parsimony of the inference
by creating a new node representing the ancestor that carried
those shared mutations and make that new node the parent of the
siblings carrying those shared mutations (Figure \ref{fig:overview_sc2ts}D). The patterns of
shared mutations between siblings can be complex, and the current
implementation uses a simple greedy strategy for choosing
the particular mutations to collapse.
